---
title: "Week 4 Assignment - HAR Predictions"
author: "Nick"
date: "12/11/2019"
output: html_document
---


## Executive Summary

This assignment models the Human Activity Recognition (HAR) data for Weight Lifting Exercises to predict the class of exercise.

The following methods were used to investigate and clean the data set including:

1. Identifying and processing missing values using by selecting a NA value threshold of 95%
2. Removing highly correlated values using the `findCorrelation` function.

Imputation was not required since the NA threshold of 95% removed all missing values in the data.

Three models were tested:

1. Gradient Boosting using the `caret` package
2. Random Forest using the `randomForest` package
3. Linear Discriminant Analysis using the `caret` package.

Of the three modelling techniques, the Random Forrest approach provided the highest accuracy and was used for the final graded assessment. The random forest model developed with the `randomForest` package performed the best overall with the lowest Out-of-Sample error (equal to 1-Accuracy) using K-fold Cross Validation.

Parallel processing techniques using the `doParallel` package were adopted to speed up computations which proved particularly useful for Gradient Boosting methods.

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively.

These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health or to find patterns in their behavior. 

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. Six study participants were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

The goal of this project is to analyse data from the accelerometers on the belt, forearm, arm, and dumbell of 6 participants and predict the class of exercise being undertaken.

More information relating to the study is available [here](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har).

Further information is available in the technical paper:
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.


## Parallel Processing
Following a Coursera Mentor's post at the link below, parallel processing was enabled to significantly improve processing performance during the model building stage.

Link to tutorial: https://github.com/lgreski/datasciencectacontent/blob/master/markdown/pml-randomForestPerformance.md.

```{r ParProc, message=FALSE, warning=FALSE}
library(parallel)
library(doParallel)
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
```

## Import Data

```{r RunLibs, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(caret)
library(e1071)
library(kableExtra)
library(randomForest)
```


```{r GetData}

  # The data has been divided into test and training sets as follows:
  train.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  test.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  
  # To improve the speed of rmarkdown, check if the data has already been loaded into the env:
  if (!exists("train.df")) {
    train.df <- read.csv(url(train.url), stringsAsFactors = FALSE)
  }
  if (!exists("test.df")) {
    test.df <- read.csv(url(test.url), stringsAsFactors = FALSE)
  }
  
```

# Data Pre-Processing

```{r}
data.dims <- data.frame(Dataset = c("train", "test"),
                        Observations = c(dim(train.df)[1], dim(test.df)[1]),
                        Variables = c(dim(train.df)[2], dim(test.df)[2]))
```

The following table shows the number of variables and observations as dimensions for each of the data sets:

```{r ShowNDims1}
kable(data.dims) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r MakeFactor}
# Ensure the dependant variable is a factor for modelling
train.df$classe <- as.factor(train.df$classe)
```

### Variable Formatting

There are `r length(which(unlist(lapply(train.df, is.character))==TRUE))` columns that are character formatted. We will now investigate these.

```{r CheckCharacterTrain, warning=FALSE}
# Identify character columns
char.col <- data.frame((train.df[, (which(unlist(lapply(train.df, is.character))==TRUE))]))
num.col <- data.frame((train.df[, (which(unlist(lapply(train.df, is.numeric))==TRUE))]))
# Show the variable names for character formatted columns
names(data.frame((train.df[, (which(unlist(lapply(train.df, is.character))==TRUE))])))
```
Next, the character variables are formatted to numeric except for the first three which should remain character formatted. The data is then checked for completeness.
```{r message=FALSE, warning=FALSE}
# Remove the Name and Dates variables.
char.col <- char.col[, -c(1:3)]

# Set the values to numeric
for (i in 1:ncol(char.col)) {
  char.col[, i] <- as.numeric(char.col[, i])
}
# round(sum(is.na(char.col))/prod(dim(char.col)), 2)
na_val_col <- sapply(char.col, 
                     function(x) mean(is.na(x))) > 0.95
```
All of the newly formatted columns contain >95% NA values and can therefore be excluded from the training and testing data set. Imputing these values would not be an accurate approach in this case.

```{r MergeData}
# Remove columns 1-7 and the previously identified columns: 
train.df <- train.df %>%
  select(-c(1:7, names(char.col)))

train.names <- train.df %>%
  select(-classe)
# Ensure the test set matches the training. The difference is the missing 'classe' dependant variable which is replaced by 'problem_id'.
test.df <- test.df %>%
  select(c(names(train.names)), problem_id)

data.dims <- data.frame(Dataset = c("train", "test"),
                        Observations = c(dim(train.df)[1], dim(test.df)[1]),
                        Variables = c(dim(train.df)[2], dim(test.df)[2]))


```

```{r ShowNDims2}
kable(data.dims) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Missing Values
The remaining data set is assessed for completeness in this section.

```{r CheckNA}
# Determine ratio of NA values
na_val_col <- sapply(train.df, 
                     function(x) mean(is.na(x))) > 0.95

na.vals <- names(which(na_val_col==TRUE))

round(sum(is.na(train.df))/prod(dim(train.df)), 2)
```

There are `r round(sum(is.na(train.df))/prod(dim(train.df)), 2) * 100`% NA values in the data set. There are `r length(na.vals)` variables with >=95% NA values and are listed in the following table.

```{r ShowNA}
na.vals
```

```{r RemoveNA}

# remove variables with > 95% NA values
na_val_col <- sapply(train.df, 
                     function(x) mean(is.na(x))) > 0.95
# Remove NA
train.df <- train.df[, na_val_col == FALSE] # Apply the filter
test.df <- test.df[, na_val_col == FALSE]

data.dims <- data.frame(Dataset = c("train", "test"),
                        Observations = c(dim(train.df)[1], dim(test.df)[1]),
                        Variables = c(dim(train.df)[2], dim(test.df)[2]))

round(sum(is.na(train.df))/prod(dim(train.df)), 2)
```

Once the high proportion NA values are removed there are no further missing values.

```{r ShowNDims3}
kable(data.dims) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Identify Correlations
Predictor correlation is assessed using the original data.

```{r Corr1}
  # Remove the dependant/outcome variable "Classe" in last column (=ncol).
  descrCor <-  cor(na.omit(train.df[, -c(ncol(train.df))])) 
  summary(descrCor[upper.tri(descrCor)])
```

The existing data has a correlation range of between `r round(summary(descrCor[upper.tri(descrCor)])['Min.'], 2)` and `r round(summary(descrCor[upper.tri(descrCor)])['Max.'], 2)`. In most cases, the strong correlations relate to the standard deviation observations of the sensors which makes sense.

These predictors can be removed using a threshold correlation cutoff of 0.75. 

```{r Corr2}
  # Find higly correlated predictors stronger than 0.75
  highlyCorDescr <- findCorrelation(descrCor, cutoff = .75) 
  # remove the highly correlated predictors. The Independant varaible is also removed as it is not numeric
  train.df.cor <- train.df[,-c(highlyCorDescr, ncol(train.df))] 
  descrCor2 <- cor(na.omit(train.df.cor))
  summary(descrCor2[upper.tri(descrCor2)])
  
  train.df.cor$classe <- train.df$classe # Add the dependant variable back in
  train.df <- train.df.cor
```

The revised correlation are now between `r round(summary(descrCor2[upper.tri(descrCor2)])['Min.'], 2)` and `r round(summary(descrCor2[upper.tri(descrCor2)])['Max.'], 2)`.

## Modelling

### Split the Data

```{r Partitioning}
# Create test and train sets since the original 'test' set is missing the Classe Variable, we cannot do cross validation.
set.seed(3456)

trainIndex <- createDataPartition(train.df$classe, p = .8, 
                                  list = FALSE, 
                                  times = 1)

Train <- train.df[ trainIndex,]
Test  <- train.df[-trainIndex,]

```

### Model Parameters for Caret Models
All of the caret models were built using the following parameters. In this case K-fold Cross Validation was used.
```{r Parameters}
fitControl <- trainControl(## 10-fold CV
                           method = "cv",
                           number = 10,
                           allowParallel = TRUE)
```

### Linear Discriminant Analysis (LDA) - CARET Package

```{r LDAModel}
set.seed(825)
mdl.lda <- train(classe ~ ., data = Train, 
                 method = "lda", 
                 trControl = fitControl,
                 verbose = FALSE)
mdl.lda
```
#### Cross Validation
```{r}
pred.lda <- predict(mdl.lda, Test)

pred.lda.cm <- confusionMatrix(pred.lda, Test$classe)

df.metrics.lda <- data.frame(Model = "Linear Discriminant Analysis", 
                             Accuracy = round(pred.lda.cm$overall['Accuracy'], 3), 
                             oos.error = 1 - round(pred.lda.cm$overall['Accuracy'], 3))
```


### Gradient Boosting - CARET Package

```{r gmbModel}
set.seed(825)
mdl.gbm <- train(classe ~ ., data = Train, 
                 method = "gbm", 
                 trControl = fitControl,
                 verbose = FALSE)
mdl.gbm
```
#### Cross Validation
```{r}
pred.gbm <- predict(mdl.gbm, Test)

pred.gbm.cm <- confusionMatrix(pred.gbm, Test$classe)

df.metrics.gbm <- data.frame(Model = "Gradient Boosting", 
                             Accuracy = round(pred.gbm.cm$overall['Accuracy'], 3), 
                             oos.error =  1 - round(pred.gbm.cm$overall['Accuracy'], 3))

```

### Random Forest - randomForest Package

```{r randomForestModel}
# Create a Random Forest model with default parameters
mdl.random.for <- randomForest::randomForest(classe ~ ., data = Train)
mdl.random.for
```
#### Cross Validation
```{r}
pred.rf <- predict(mdl.random.for, Test, type = "class")

pred.randForr.cm <- confusionMatrix(pred.rf, Test$classe)

df.metrics.rand <- data.frame(Model = "Random Forest", 
                              Accuracy = round(pred.randForr.cm$overall['Accuracy'], 3), 
                              oos.error = 1- round(pred.randForr.cm$overall['Accuracy'], 3))
```

## Compare Models - Out of Sample Accuracy
The cross validated accuracies of each of the models is compared in the following table to provide the out of sample accuracies:

```{r Comparison}
df.comparison <- rbind(df.metrics.gbm, df.metrics.rand, df.metrics.lda)
row.names(df.comparison) <- NULL
kable(df.comparison %>%
  arrange(desc(Accuracy))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


Based on the above table the Random Forrest model from the `randomForest` package yeilds the highest out of sample accuracy and will be used in the final predictions for the quiz.
```{r}
# Export the final model for later reuse and time saving
saveRDS(mdl.random.for, "randomForest_Final.rds")
```

## De-register Parallel Processing Cluster

```{r}
stopCluster(cluster)
registerDoSEQ()
```

## Final Quiz
The Random Forrest model generated using the `randomForest` package is used to predict the Class of exercise using the original testing data. These predictions will then be used answer the final quiz. 

The following table shows the prediction against the Problem ID.

```{r}
# Read in the final model
mdl.random.for <- readRDS("randomForest_Final.rds")

pred.rf.final <- predict(mdl.random.for, test.df)
# Checking classification accuracy

test.predictions <- data.frame(Problem_id = 1:20,
                               Prediction = pred.rf.final
                               )
row.names(test.predictions) <- NULL

kable(test.predictions) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

